---
title: "Bootstrapping"
author: "Stefan Eng"
date: "5/26/2021"
output:
  beamer_presentation:
    slide_level: 2
    md_extensions: +grid_tables
  html_document: default
bibliography: references.bib
---

```{r, setup, echo=FALSE, warning = FALSE, message = FALSE, results = 'hide'}
library(boot)
library(knitr)
library(kableExtra)
library(BoutrosLab.plotting.general)

set.seed(13)

knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, fig.align = 'center', out.width = '90%', fig.pos = 'H');

# CB Palette like ESL
cbPalette <- c(
  "#999999", "#E69F00", "#56B4E9", "#009E73",
  "#F0E442", "#0072B2", "#D55E00", "#CC79A7")
```

## Overview
  - Bootstrapping provides a way to quantify uncertainty of an estimator
  - The basic idea is that we resample, with replacement, from our sample and use the distribution of those resamples to compute the standard error and confidence intervals.

## Random Variable
  - A random variable is a mapping (a function) $X : \Omega \to \mathbb R$ that assigns a real number $X(\omega)$ to each outcome $\omega$.
  - We almost never write $X(\Omega)$ but simply write $X$
  - Example: Flip a coin and let $X$ be the number of heads shown
  - The sample space is $\Omega = \{\{ H, H\}, \{ H, T\}, \{ T, H\}, \{T, T\} \}$

```{r}
coin.flip <- data.frame(coin1 = c('H', 'H', 'T', 'T'), coin2 = c('H', 'T', 'H', 'T'))
coin.flip$numHeads <- rowSums(coin.flip == 'H')
kable(coin.flip, col.names = c('coin 1', 'coin 2', 'Number of heads'), row.names = FALSE)
```  
  
## Random Variable - Dice Example
  - The sample space is $\Omega = \{\{ 1, 1\}, \{ 1, 2\}, \{ 2, 1\}, \ldots \}$
  - $X(\{3,4\}) = 7$

```{r}
dice <- data.frame(die1 = unlist(lapply(1:6, rep, times = 6)), die2 = rep(1:6, 6))
dice$sum <- rowSums(dice)
dice <- dice[order(dice$sum), ]
kable(head(dice), row.names = FALSE, col.names = c('die 1', 'die 2', 'Sum of dice'))
```

## Cumulative Distribution Function
  For a random variable $X$, the **cumulative distribution function** (CDF), $F_X : \mathbb R \to [0,1]$ is
  $$
  F_X(x) = P(X \leq x)
  $$

## Notation
  - If $X \sim F$ then we say that $X$ has distribution $F$.
  - For example, $X \sim \exp(\lambda)$ means that $X$ is exponentially distributed with rate $\lambda$ and
  $$
  F(x) = \begin{cases}
    1 - e^{-\lambda x} & x \geq 0\\
    0 & x < 0
  \end{cases}
  $$

## CDF - Normal
```{r}
x <- seq(-4, 4, length.out= 100)
plot(x, pnorm(x), type = "l", ylab = "F(x)", main = "Normal distribution CDF", las = 1)
lines(x, pnorm(x, sd = 1/2), lty = 2)
lines(x, pnorm(x, sd = 2), lty = 3)
legend(2, 0.4,
       legend=c(
         expression(paste(sigma, "=", 1)),
         expression(paste(sigma, "=", 0.5)),
          expression(paste(sigma, "=", 2))),
       lty=1:3, cex=0.8)
```

## Empirical Distribution Function
  - Let $X_1, \ldots, X_n \sim F$ be iid random variables
  - Note that this means that $X$ has a distribution function $F$.
  - We can estimate $F$ from the data by using the empirical distribution function $\hat{F}_n$.
  - This distribution $\hat{F}_n$ puts probability of 1/n on each data point

## Example - Normal Distribution
```{r}
set.seed(10)
norm.sample <- rnorm(100)
hist(norm.sample)
```

## Example - Normal (E)CDF
```{r}
N <- c(15, 200)
set.seed(13)
x <- rnorm(N[1])
x.range <- seq(-3, 3, length.out = 1000)
x.F <- pnorm(x.range)
x.ecdf <- ecdf(x)

par(mfrow=c(1,2), mai = c(1, 0.2, 0.5, 0), oma = c(0, 1.2, 0.5, 0.5), las = 1)
plot(ecdf(x), verticals = TRUE, xlim = c(-3, 3), main = '')
rug(x)
lines(x.range, x.F, lty=2)
axis(side = 2, seq(0, 1, by = 0.2))

x <- rnorm(N[2])
x.range <- seq(-3, 3, length.out = 1000)
x.F <- pnorm(x.range)
x.ecdf <- ecdf(x)
plot(ecdf(x), verticals = TRUE, xlim = c(-3, 3), ylab = '', yaxt="n", main = '')
rug(x)
lines(x.range, x.F, lty=2)
title(sprintf('(E)CDF of Normal Distribution n = %d, n = %d', N[1], N[2]), outer = TRUE, line = -1)
```

## Sampling Distribution
  - Let $X_1, \ldots, X_n$  be iid random variables
  - A function of the data is called a **statistic**
  - The mean of these random variables is an example of a statistic
  $$
  \overline{X}_n = \frac{1}{n} \sum_{i = 1}^n X_i
  $$
  - $\overline{X}_n$ is itself a random variable, and thus has a distribution (called the sampling distribution of the statistic).
  - Example: If $X_1, \ldots, X_n \sim N(\mu,\sigma)$ then $\overline{X}_n \sim N \left(\mu, \frac{\sigma}{\sqrt{n}} \right)$

## Statistical Inference
  - Given a sample $X_1, \ldots, X_n \sim F$ we want to infer the distribution $F$.
  - We approximate $F$ using a statistical model, which is a set of distributions
    - Parameteric models use a set of distributions that can be parameterized by a finite set of parameters.
    - Example: Two-parameter model for a set of Gaussians
    - Non-parametric models cannot be parameterized by a finite set of parameters

## Example
  - Say we have a sample of iid random variables $X_1, \ldots, X_n$.
  - Computing the variance (and confidence intervals) of the mean is relatively easy.
  - What about some arbitrary statistic: $T_n = g(X_1, \ldots, X_n)$?
    - Option 1: Do some possibly complicated mathematics.
    - Option 2: Bootstrap

## Bootstrap
> "The population is to the sample as the sample is to the bootstrap samples." (Fox 2008)

  1. Compute $T_n = g(X_1, \ldots, X_n)$, our statistic of interest.
  2. Draw a sample $X_1^*, \ldots, X_n^* \sim \hat{F}_n$
    - All this means is to sample $n$ times with replacement from the original data $X_1, \ldots, X_n$.
  3. Compute our statistic of interest $T_n^* = g(X_1^*, \ldots, X_n^*)$
  4. Repeat $B$ times, to get $T_{n,1}^*, \ldots, T_{n,B}^*$
  5. Compute the variance of $T_{n,1}^*, \ldots, T_{n,B}^*$ to get $v_{\text{boot}}$, and standard error $\hat{\text{se}}_{\text{boot}} = \sqrt{v_{\text{boot}}}$
  
## Bootstrap
  - Bootstraping does not improve our original estimate in any way
  - The idea is to provide a way of estimating the uncertainty in the computed statistic
  - Need to have a fairly large sample to get accurate estimation, especially if the statistic depends on a small number of observations (like the median).
  - Highly skewed distributions may not work well for bootstrapping without a transformation
  
## Approximations
$$
\operatorname{Var}_F(T_n)  \approx \operatorname{Var}_{\hat{F}_n} (T_n) \approx v_{\text{boot}}
$$

  - Multiple approximations are happening during bootstrapping (different sources of error)
  - First we approximate $F$ with $\hat{F}_n$. Error depends on how big the sample is.
  - Approximating $\operatorname{Var}_{\hat{F}_n} (T_n)$ by $v_{\text{boot}}$ depends on the size of the bootstrap samples $B$.

## Bias
The bias of an estimator $\hat{\theta}$ is
$$
\operatorname{B} = E(\hat \theta) - \theta
$$
We can estimate the bias with 
$$
\operatorname{B} = E_{\hat F}(\hat{\theta^*}) - \hat \theta
$$
That is, the difference between the mean of the bootstrap distribution and the observed statistic.

## Bootstrap Confidence Intervals
  - Once we have $\hat{\text{se}}_{\text{boot}}$ how do we compute a confidence interval?

  - **Normal** Interval: Don't use unless distribution of $T_n$ is close to normal.
  $$
  T_n \pm z_{\alpha/2} \hat{\text{se}}_{\text{boot}}
  $$

  - **Percentile** Interval: Simply use the $\alpha / 2$ and $1 - \alpha / 2$ quantiles of the bootstrap sample.
    - For small samples, may not be accurate
    - Benefit is that they are transformation invariants, you can apply a monotone transformation to the data and get the same CI after inverse transformation.
    - In general the following methods are more accurate though
    
## Bootstrap Confidence Intervals
  - **Basic** (Pivotal) Interval: Incorporate the bias into the confidence interval.
  - **Studentized** Interval: Need to compute the standard error of each of the bootstrap samples
  - **Bias Corrected, Accelerated (Bca)**: estimate a bias and acceleration term.
    - Corrects for skew in sampling distribution.
    - Requires a large number of bootstrap samples

## LSAT Scores and GPAs (Efron and Tibshirani 1993)
An example dataset with students' LSAT and GPAs (n = 15). Shown are the first 5 students.
```{r}
lsat.gpa <- read.table(text = "student	lsat	gpa
1	576	3.39
2	635	3.30
3	558	2.81
4	578	3.03
5	666	3.44
6	580	3.07
7	555	3.00
8	661	3.43
9	651	3.36
10	605	3.13
11	653	3.12
12	575	2.74
13	545	2.76
14	572	2.88
15	594	2.96", header = TRUE)
kable(head(lsat.gpa, n = 5))
```

## Bootstrap Example
  - `mtcars` built-in data set in R. 
  - Correlation of car weight and miles per gallon (mpg)
```{r, echo = TRUE, eval = FALSE}
library(boot)
boot.cor <- function(data, indices) {
  d <- data[indices,]
  cor(d$wt, d$mpg, method = "spearman")
}

results <- boot(data=mtcars, statistic=boot.cor, R=1000)
results
```

## Bootstrap Example
```{r}
boot.cor <- function(data, indices) {
  d <- data[indices,]
  cor(d$wt, d$mpg, method = "spearman")
}

results <- boot(data=mtcars, statistic=boot.cor, R=1000)
results
```

## Bootstrap Example
```{r}
plot(results)
```

## Bootstrap Confidence Intervals
\small
```{r, echo = TRUE}
boot.ci(results, type = c("basic", "bca"))
```
\normalsize

## Bootstrap Confidence Intervals
```{r}
boot.R2 <- function(data, indices) {
  d <- data[indices,]
  summary(lm(mpg ~ wt + disp, data = d))$r.square
}

results.R2 <- boot(data=mtcars, statistic=boot.R2, R=1000)
plot(results.R2)
```

## Bootstrap Confidence Interval
```{r}
boot.ci(results.R2, type = c("basic", "bca"))
```

## Bootstrap Linear Models
First we look at the estimates for `mpg ~ wt + disp`
```{r, echo = TRUE}
mpg.mod <- lm(mpg ~ wt + disp, data = mtcars)
confint(mpg.mod)
```

## Bootstrap Linear Models
```{r, echo = TRUE}
boot.lm <- function(formula, data, indices) {
  d <- data[indices,]
  coef(lm(formula, data = d))
}
results.lm <- boot(data=mtcars, statistic=boot.lm, R=1000, formula = mpg ~ wt + disp)

# Confidence interval for wt coefficient
boot.ci(results.lm, type = c("basic", "bca"), index = 2)
```

## Other version of the bootstrap
What we have been doing is the nonparameteric bootstrap.
 Other options include
 
 - **Semiparametric bootstrap**: Add noise to the resamples to produce non-identical resamples
 - **Parametric bootstrap**: Assume the data comes from a known distribution and estimate the parameters given the data. Use this estimated distribution to draw samples.
 - **Block bootstrap**: When the data is no longer iid, and correlations between data or errors exists.
   - For example, bootstrap on time-series data
   - See `boot::tsboot`

## Permutation Tests
  
```{r, eval = FALSE}
x <- data.frame(x = rnorm(30, mean = -2, sd = 4), group = 1)
y <- data.frame(x = rnorm(30, mean = 2, sd = 0.5), group = 2)
data <- rbind(x, y)
data$group <- as.factor(data$group)
create.boxplot(
  formula =  x ~ group,
  data = data
)

mean(x$x) - mean(y$x)

res <- replicate(1000, {
  s <- sample(data$x)
  mean(s[1:30]) - mean(s[31:60])
})

hist(res)
```

## Wilcoxon, Mann-Whitney
https://blog.minitab.com/en/adventures-in-statistics-2/choosing-between-a-nonparametric-test-and-a-parametric-test

## Smoothing
...

## References
  - [Bootstrap confidence intervals](https://blog.methodsconsultants.com/posts/understanding-bootstrap-confidence-interval-output-from-the-r-boot-package/)
  - [Permutation Tests](https://faculty.washington.edu/kenrice/sisg/SISG-08-06.pdf)
  - [What Teachers Should Know About the Bootstrap: Resampling in the Undergraduate Statistics Curriculum](https://www.tandfonline.com/doi/full/10.1080/00031305.2015.1089789)
